---
title: "exercise3"
author: "Elsie Backen Tandberg, Maja Mathiassen, Florian Beiser"
date: "23.04.2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(MASS)
```

# Problem B

```{r}
bilirubin <- read.table("https://www.math.ntnu.no/emner/TMA4300/2020v/Exercise/ex3-additionalFiles/bilirubin.txt",header=T)
bilirubin

```

## B1
```{r}
#Finding the boxplot 
boxplot(log(meas)~pers, data=bilirubin)
```


```{r}
#Fitting a linear model to the data
mod <- lm(log(meas)~pers, data=bilirubin)
#Finding the F-statistic
summ <- summary(mod)
#Saving the value of the F statistic as Fval
Fval <- summ$fstatistic[1]
#Calculating the quantlie function
qf(0.95, 2, 26)
```

We will use the F-test to test the hypothesis that 
$$
H_0: \beta_1=\beta_2=\beta_3 \quad \text{vs}\quad H_1:\text{at least one is different.}
$$


The F-statistic was found from the summary of the fitted linear model. $H_0$ is rejected if the F-statistic = `Fval` is higher than `r qf(0.95, 2, 26)`. We see that this is true, so we reject $H_0$, concluding that there at least one of the $\beta$'s differ from the others.

## B2
Now we will write a function that generates a permutation of the data between the three individuals. 
```{r}
permTest <- function(){
  #Generating one permutation of the dataset
  perm <- data.frame(meas=sample(bilirubin$meas, 29), pers=bilirubin$pers)
  #Fitting a linear model to the data
  mod <- lm(log(meas)~pers, data=perm)
  #Finding the F-statistic
  summ <- summary(mod)
  #Saving the value of the F statistic as Fval
  Fval <- summ$fstatistic[1]
  return(Fval)
}
```

## B3
We will generate 999 samples of the F-value by using the permutations calculated by the function permTest from B2. Theese samples will be used to find the p-value by counting the number of observations that are greater than the F-value found in B1. This p-value will be used to evaluate the same hypothesis test as in B1, with significance level $\alpha=0.05$.
```{r}
#Generating 999 samples of the F-value from the permTest function
samples <- rep(0,999)
for(i in 1:999){
  samples[i] <- permTest()
}
#Finding the p-value
pval <- sum(samples > Fval)/999
```
We found our p-value to be `r pval`, this is lower than our significance level, meaning that we reject $H_0$. So the same conclusion as in B1 is reached, where we believe that the three individuals have a differing concentration of bilirubin in their blood.


# Problem C
Here we consider $x_1, ..., x_n$ and  $y_1, ..., y_n$ which are exponential distributed independent random variables with with intensity $\lambda_0$ and $\lambda_1$ respectively. We do not observe $x_1, ..., x_n$ and  $y_1, ..., y_n$ directly, instead we observe $z_i=\max(x_i, y_i)$ for $i=1,...,n$, and $u_i=I(x_i\ge y_i)$ for $i=1,...,n$. Here $I(A)=1$ if $A$ is true, and $0$ otherwise. Based on the observations $(z_i, u_i)$ we will use the EM algorithm to find the maximum likelihood estimates for $(\lambda_0, \lambda_1)$.

## C1
First the log likelihood function for the complete data $(x_i, y_i)$, $i=1,...,n$ is found.

$$
\begin{aligned}
L(x,y)&=\prod_{i=1}^n \lambda_0 e^{-\lambda_0x_i}\lambda_1 e^{-\lambda_1y_i}\\
l(x,y)&= n(\log(\lambda_0)+\log(\lambda_1))-\lambda_0\sum_{i=1}^n x_i-\lambda_1\sum_{i=1}^n y_i.
\end{aligned}
$$
Further we are interested in the expectation of the log likelihood
$$
E[\log(f(\mathbf{x}, \mathbf{y}|\lambda_0, \lambda_1))|\mathbf{z}, \mathbf{u}, \lambda_0^{(t)}, \lambda_1^{(t)})]=n(\log(\lambda_0)+\log(\lambda_1))-\lambda_0\sum_{i=1}^n E(x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})-\lambda_1\sum_{i=1}^n E(y_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)}).
$$
The expressions for $E(x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})$ and $E(y_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})$ are dependent on which of the variables $(x_i, y_i)$ we have observed. So the pdf´s $f((x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)}))$ and $f((y_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)}))$ can be written as two split equations depending on the value of $u_i$.

$$
\begin{aligned}
f(x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})&=\begin{cases}
z_i & u_i=1\\
\frac{f(z_i|x_i)f(x_i)}{f(z_i)} & u_i=0\\
\end{cases}\\
&=\begin{cases}z_i & u_i=1\\
\frac{\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}}{\int_0^{z_i}\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}dx_i}& u_i=0\\
\end{cases}\\
&=\begin{cases}z_i & u_i=1\\
\frac{\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}}{1-e^{-\lambda_0^{(t)}z_i}}& u_i=0\\
\end{cases}
\end{aligned}
$$


$$
E(x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})=z_iu_i+(1-u_i)\frac{\int_0^{z_i}x_i\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}dx_i}{1-e^{-\lambda_0^{(t)}x_i}}
$$

The integral $\int_0^{z_i}x_i\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}dx_i$ is found by integration by parts.

$$
\begin{aligned}
\int_0^{z_i}x_i\lambda_0^{(t)}e^{-\lambda_0^{(t)} x_i}dx_i&=-x_ie^{\lambda_0^{(t)}x_i}\bigg|_0^{z_i}-\int_0^{z_i}-e^{-\lambda_0^{(t)}x_i}dx_i\\
&=-z_ie^{\lambda_0^{(t)}z_i}-\bigg[\frac{1}{\lambda_0}e^{-\lambda_0^{(t)}x_i}\bigg]_0^{z_i}\\
&=\frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1}
\end{aligned}
$$
Inserting this into the expression for the expectation we get the following equation
$$
E(x_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})=z_iu_i+(1-u_i)\bigg(\frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1}\bigg)
$$
The same calculations are done to find $E(y_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})$

$$
E(y_i|z_i, u_i, \lambda_0^{(t)}, \lambda_1^{(t)})=(1-u_i)z_i+u_i\bigg(\frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1}\bigg)
$$

Combining the exressions for the expecatation leads to the following equation

$$
\begin{aligned}
E[\log(f(\mathbf{x}, \mathbf{y}|\lambda_0, \lambda_1))|\mathbf{z}, \mathbf{u}, \lambda_0^{(t)}, \lambda_1^{(t)})]&=n(\log(\lambda_0)+\log(\lambda_1))\\
&-\lambda_0\bigg[z_iu_i+(1-u_i)\bigg(\frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1}\bigg)\bigg]\\
&-\lambda_1\bigg[(1-u_i)z_i+u_i\bigg(\frac{1}{\lambda_0^{(t)}}-\frac{z_i}{e^{\lambda_0^{(t)}z_i}-1}\bigg)\bigg].
\end{aligned}
$$





## C2
```{r}
u <- read.table("https://www.math.ntnu.no/emner/TMA4300/2020v/Exercise/ex3-additionalFiles/u.txt",header=F)
z <- read.table("https://www.math.ntnu.no/emner/TMA4300/2020v/Exercise/ex3-additionalFiles/z.txt",header=F)
```


```{r}
#EM algorithm 
EM.algorihtm <- function(lam0, lam1, u, z){
  lam0i <- lam0
  lam1i <- lam1
  n<-length(u)
  diff <- 1
  lambda0.list <- c(lam0i)
  lambda1.list <- c(lam1i)
  
  while(diff>1e-10){
    lamvec <- c(lam0i, lam1i)
    #Expectation step
    sum0 <- sum(u*z+(1-u)*(1/lam0i-z/(exp(lam0i*z)-1)))
    sum1 <- sum((1-u)*z+u*(1/lam1i-z/(exp(lam1i*z)-1)))
    #Q <- n*(log(lam0i)+log(lam1i))-lam0i*sum0-lam1i*sum1
    
    #Maximization step
    lam0i <- n/sum0
    lam1i <- n/sum1
    
    #Storing the lambda values
    lambda0.list <- c(lambda0.list, lam0i)
    lambda1.list <- c(lambda1.list, lam1i)
    
    #Convergence criteria
    diff <- norm(cbind(lamvec[1]-lam0i, lamvec[2]-lam1i), type="2")
    
  }
  #print(lambda0.list)
  return(cbind(c(lambda0.list), c(lambda1.list)))
}

em<-EM.algorihtm(0.5,0.4, u$V1, z$V1)
theta_hat = em[length(em[,1]), ]
theta_hat

x <- seq(1:length(em[,1]))
ggplot()+
  geom_point(aes(x=x, y=em[,1], color="lambda0"))+
  geom_point(aes(x=x, y=em[,2], color="lambda1"))  +  xlab('Iterations') + ylab(expression(lambda))  
```




``` {r}
n <- 200
B <- 1000
df <- data.frame(u = u$V1, z = z$V1)
bootstrap.lambda <- data.frame(lambda0 = rep(0, n), lambda1 = rep(0, n))

for(i in 1:B) {
  index <- sample(1:200, n, replace = TRUE)
  bootstrap.sample <- df[index, ]
  em <- EM.algorihtm(0.5, 0.5, bootstrap.sample[,1], bootstrap.sample[,2])
  bootstrap.lambda[i,] <- em[length(em[,1]),]
}

mean <- colMeans(bootstrap.lambda)
mean

sd0 <- sqrt( 1/(B-1) * sum((bootstrap.lambda[,1] - mean[1])^2) )
sd1 <- sqrt( 1/(B-1) * sum((bootstrap.lambda[,2] - mean[2])^2) )

sd0
sd1

sd(bootstrap.lambda[,1])
sd(bootstrap.lambda[,2])
```



```{r}
l0 = theta_hat[1]
l1 = theta_hat[2]
u = u$V1
z = z$V1
```

```{r}
n = length(u)
n0 = sum(u==0)
n1 = n - n0
n0; n1

logLikelihood = function(par, u, z) {
  summ = n0*log(par[2]) + n1*log(par[1])
  summ = summ + sum( u*(log(1-exp(-par[2]*z)) - par[1]*z) + (1-u)*(log(1-exp(-par[1]*z)) - par[2]*z) )
  return(-summ)
}

optim(par=c(2,2), fn=logLikelihood, u=u, z=z)$par
```











